Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-12-06 paper/manuscript.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
\author[7]{Bryce K. Allen}
\author[1*]{John D. Chodera}
\author[2,12]{Maximilian Kuhn}
\author[2*]{Antonia S. J. S. Mey}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[1]{Levi N. Naden}
\author[4*]{Samarjeet Prasad}
\author[5*]{Julia E. Rice}
\author[1,9]{Andrea Rizzi}
\author[2*]{Jenke Scheen}
\author[6*]{Michael Shirts}
\author[11]{Gary Tresadern}
\author[10*]{Huafeng Xu}

@@

@@

    \includegraphics[width=0.95\columnwidth]{paper/figures/fig2_what_is_lam/Fig2.pdf}

@@

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig3_therm_cycle/Fig3.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:fig3_binding-thermodynamic-cycle}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

\section{Drug discovery}

@@

The concerns mentioned so far, such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup, which is beyond the scope of this article, there are some additional points to consider. Formal charge, or tautomeric state of the small molecules can change within a series. The pKa can change by several log units with only small structural modifications and this is a tactic often used within a congeneric series to modify drug properties. Check up-front because charge or tautomeric change would require modified approaches. A second concern is to try to capture aspects of the bioassay into the simulations, for instance what pH was used for screening? Does the protein X-ray match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? Were certain co-factors or partner proteins required to augment activity? 
%

As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, normally solved via X-ray crystallography. If more than one structure is available, the modeler should pay attention to choose the most suitable. The quality of the structure can be a concern, and the reader is referred to work of Warren et al for a detailed discussion of choosing optimal structures for structure-based modeling (CITE: Warren et al Drug Discov Today, 2012, 17, 1270). Particularly relevant for alchemical FE calculations is to study the SAR and understand the impact in the binding site, is side chain movement required and is there evidence of this in any alternative X-ray structures of the same protein? Normally, only one protein and water configuration is used, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations. This can be a practical limitation for the extent of structural change that can be attempted within a series, a simple work around can be to separate compounds into subseries for different calculations. If multiple structures are available there is some evidence the higher affinity complex can give better performance (cite Perez-Benito jctc 2019, 15, 1884). 
%

@@

\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig4_topol/Fig4.pdf}
    \caption{top: single topology bottom: dual topology Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:fig4_topology}
\end{figure} 
The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in which functional groups are present; e.g. consider a transformation of benzene (A) to benzyl alcohol (B)~\ref{fig:topology}.
In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group. Bond parameters as well as partial charges are adjusted accordingly. 
Thus in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:fig4_topology} (a). 
In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{Shirts2012}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{Mobley:2014:J.Comput.AidedMol.Des.}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:topology} (b) depicts how such a dual topology works.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95*\columnwidth]{paper/figures/fig7_sampl_scheme/sampling_schemes.pd}
    \caption{\textbf{There are 4 different sampling strategies available and implemntations are availble for different software packages.} A: Multiple replicas in parallel at different lambda states. B: Hamiltonian replica exchange scheme. C: Single replica scheme sampling from all $\lambda$ states. D: Non-equilibrium sampling scheme.}
    \label{fig:fig7_sampling_scheme}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[most]{tcolorbox}
\newcommand{\versionnumber}{0.1}  % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}}  %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
\author[7]{Bryce K. Allen}
\author[1*]{John D. Chodera}
\author[2,12]{Maximilian Kuhn}
\author[2*]{Antonia S. J. S. Mey}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[1]{Levi N. Naden}
\author[4*]{Samarjeet Prasad}
\author[5*]{Julia E. Rice}
\author[1,9]{Andrea Rizzi}
\author[2*]{Jenke Scheen}
\author[6*]{Michael Shirts}
\author[11]{Gary Tresadern}
\author[10*]{Huafeng Xu}

@@

\begin{tcolorbox}[colback=blue!5!white,colframe=blue!75!black]
    \begin{center}
        \textit{In alphabetical order:}
    \end{center}
    \textit{CPU} - Central Processing Unit\\
    \textit{FE} - Free Energy\\
    \textit{FEP} - Free Energy Perturbation\\
    \textit{GPCR} - G-Protein Coupled Receptor\\
    \textit{GPU} - Graphics Processing Unit\\
    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} - Maximum Common SubStructure\\
    \textit{MD} - Molecular Dynamics\\
    \textit{RBFE} - Relative Binding Free Energy\\
    \textit{RMSE} - Root Mean Square Error\\
    \textit{SAR} - Structure-Activity Relationships\\
    \textit{TI} - Thermodynamic Integration

\end{tcolorbox}

@@

    \includegraphics[width=0.95\linewidth]{paper/figures/fig2_what_is_lam/Fig2.pdf}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{paper/figures/fig3_therm_cyc/Fig3.pdf}
    \caption{{\bf Thermodynamic cycle for computing the relative free energy of binding ($\Delta \Delta G$) between two related small molecules to a supramolecular host.}
    The relative binding free energy difference between two small molecules, $\Delta \Delta G_{\mathrm{bind}, A \rightarrow B} \equiv \Delta G_{\mathrm{bind}, B} - \Delta G_{\mathrm{bind}, A}$---here benzyl alcohol (top) to toluene (bottom)---can be computed as a difference between two alchemical transformations, $\Delta G_\mathrm{bound} - \Delta G_\mathrm{solvated}$, where $\Delta G_\mathrm{bound}$ represents the free energy change of transforming $A \rightarrow B$ in complex and $\Delta G_\mathrm{solvated}$ the free energy change of transforming $A \rightarrow B$ in solvent.}
    \label{fig:fig3_binding-thermodynamic-cycle}
\end{figure}
\todo[inline, color=green!20]{This section needs to be expanded and give a good introduction to the theoretical background without going into too much detail.}

@@

\section{How should alchemical simulations be applied to drug discovery?}

@@

The concerns mentioned in the previous section such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup, which is beyond the scope of this article, there are some additional points to consider. Formal charge, or tautomeric state of the small molecules can change within a series. The pKa can change by several log units with only small structural modifications and this is a tactic often used within a congeneric series to modify drug properties. Check up-front because charge or tautomeric change would require modified approaches. A second concern is to try to capture aspects of the bioassay into the simulations, for instance what pH was used for screening? Does the protein X-ray match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? Were certain co-factors or partner proteins required to augment activity? 
%
\subsection*{Is my binding mode accurate?}
As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, normally solved via X-ray crystallography. If more than one structure is available, the modeler should pay attention to choose the most suitable. The quality of the structure can be a concern, and the reader is referred to work of Warren et al for a detailed discussion of choosing optimal structures for structure-based modeling (CITE: Warren et al Drug Discov Today, 2012, 17, 1270). Particularly relevant for alchemical FE calculations is to study the SAR and understand the impact in the binding site, is side chain movement required and is there evidence of this in any alternative X-ray structures of the same protein? Normally, only one protein and water configuration is used, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations. This can be a practical limitation for the extent of structural change that can be attempted within a series, a simple work around can be to separate compounds into subseries for different calculations. If multiple structures are available there is some evidence the higher affinity complex can give better performance (cite Perez-Benito jctc 2019, 15, 1884). 
%

@@

\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.
\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig4_topol/Fig4.pdf}
    \caption{top: single topology bottom: dual topology Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}
    \label{fig:fig4_topology}
\end{figure} 
The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in which functional groups are present; e.g. consider a transformation of benzene (A) to benzyl alcohol (B)~\ref{fig:topology}.
In this case the common substructure is the benzene ring, though perhaps may be larger depending on how it is defined, as we discuss below.
In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group. Bond parameters as well as partial charges are adjusted accordingly. 
Thus in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in fig.~\ref{fig:fig4_topology} (a). 
In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{Shirts2012}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{Mobley:2014:J.Comput.AidedMol.Des.}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:topology} (b) depicts how such a dual topology works.

@@

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{paper/figures/fig7_sampl_scheme/sampling_schemes.pdf}
    \caption{\textbf{There are 4 different sampling strategies available and implemntations are availble for different software packages.} A: Multiple replicas in parallel at different lambda states. B: Hamiltonian replica exchange scheme. C: Single replica scheme sampling from all $\lambda$ states. D: Non-equilibrium sampling scheme.}
    \label{fig:fig7_sampling_scheme}
\end{figure}
